Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data

### 2019-05-31 thesis.tex
Remove formatting specific to the thesis submission

\documentclass[12pt]{article}
\usepackage{titling}

\title{Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data}
\author{Jonathan Bryan Dayton}
% \author[1]{Jonathan B. Dayton}
% \author[1]{Stephen R. Piccolo}
% \affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}
\date{2019}

% Times New Roman for the thesis submission
% \usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{authblk}
\usepackage{url}
\usepackage[inline,shortlabels]{enumitem}
\usepackage[square,numbers]{natbib}
% \usepackage{natbib}
\usepackage{floatrow}
\floatsetup[table]{capposition=top}
\usepackage[nomarkers,nolists,tabhead,fighead,tablesfirst]{endfloat}
% \renewcommand{\efloatseparator}{\mbox{}} % multiple figures per page
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{csvsimple}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{changepage}
\usepackage{tocloft}
\usepackage[bookmarksnumbered,bookmarksopen=true,bookmarks=true,hidelinks,pdfpagelabels=true]{hyperref}
\usepackage{bookmark}

% Make all figures and tables hang towards the top
\makeatletter
  \providecommand*\setfloatlocations[2]{\@namedef{fps@#1}{#2}}
\makeatother
\setfloatlocations{figure}{htbp}
\setfloatlocations{table}{htbp}

% Prevent captions from setting stuff 2x and therefore making 2 pdfbookmarks
\usepackage{caption}
\captionsetup{singlelinecheck=off}

% Some manual \pdfbookmarks are too low. This should raise them.
\makeatletter
 \newcommand{\pdfbookmarkraised}[3]{\Hy@raisedlink{\pdfbookmark[#1]{#2}{#3}}}
\makeatother

% TODO: change `\nolinkurl`s to `\url`s
% Add dots to sections in TOC
\renewcommand{\cftsecleader}{\cftdotfill{\cftdotsep}}

% stuff for restyling the table of contents
\renewcommand\cftsecfont{\mdseries}
\renewcommand\cftsecpagefont{\mdseries}
\renewcommand{\contentsname}{\begin{center}\normalsize{\textnormal{TABLE OF CONTENTS}}\end{center}}
\renewcommand{\listtablename}{\begin{center}\normalsize{\textnormal{LIST OF TABLES}}\end{center}}
\renewcommand{\listfigurename}{\begin{center}\normalsize{\textnormal{LIST OF FIGURES}}\end{center}}

@@

\pagenumbering{roman}
\begin{titlingpage}
	\setcounter{page}{1}
	\addcontentsline{toc}{section}{Title Page}
	\begin{adjustwidth}{+0.5in}{+0.5in}
		\begin{center}
			\doublespacing \thetitle\\ \singlespacing
		\end{center}
	\end{adjustwidth}
	\centering{
		\vfill
		\theauthor\\
		\vfill
		A thesis submitted to the faculty of\\
		Brigham Young University\\
		in partial fulfillment of the requirements for the degree of\\
		\vspace{1em}
		Master of Science\\
		\vfill
		Stephen R. Piccolo, Chair\\
		Mark Clement\\
		David Wingate\\
		\vfill
		Department of Biology\\
		\vspace{1em}
		Brigham Young University
		\vfill
		Copyright \textcopyright \ \thedate \ \theauthor\\
		\vspace{1em}
		All Rights Reserved
	\par}

\end{titlingpage}

% Title / abstract page
\begin{titlingpage}
	\setcounter{page}{2}
	\phantomsection
	\addcontentsline{toc}{section}{Abstract}
	\begin{center}
		ABSTRACT\\
		\vspace{1em}
		\begin{adjustwidth}{+0.5in}{+0.5in}
			\begin{center}
				\thetitle\\
			\end{center}
		\end{adjustwidth}
		\vspace{1em}
		\theauthor\\
		Department of Biology, BYU\\
		Master of Science
	\end{center}
	% \begin{abstract}
		\paragraph{} % adds indentation
		Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
		However, because of the high dimensionality of these data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
		For example, collecting data in multiple runs causes nontrivial differences in the data (known as batch effects),
		known covariates that are not of interest to the study may have strong effects,
		and there may be large systemic effects when integrating multiple expression datasets.
		Additionally, many of these confounding effects represent higher-order interactions that may not be removable using existing techniques that identify linear patterns.
		We created Confounded to remove these effects from expression data.
		Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
		We tested the model on artificially constructed data and commonly used gene expression datasets and compared against other common batch adjustment algorithms.
		We also applied the model to remove cancer-type-specific signal from a pan-cancer expression dataset.
		Our software is publicly available at \nolinkurl{https://github.com/jdayton3/Confounded}.
	% \end{abstract}
	\vfill
	\noindent
	Keywords:\@ batch effects, batch correction, gene expression, transcriptomics, deep learning, adversarial neural network, variational autoencoder % the \@ removes the double "\nonfrenchspacing" spaces after the colon, which is required for thesis submission
\end{titlingpage}

\doublespacing

\begin{titlingpage}
	\setcounter{page}{3}
	\phantomsection
	\addcontentsline{toc}{section}{Acknowledgments}
	\begin{center}
		ACKNOWLEDGMENTS\\
		TODO: Fill this out.
	\end{center}
\end{titlingpage}

\setcounter{page}{4}
\phantomsection
\addcontentsline{toc}{section}{Table of Contents}
\tableofcontents
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of \tablename{s}}
{% Change table refs "1", "2" -> "Table 1", "Table 2" in LOT
\let\oldnumberline\numberline%
\renewcommand{\numberline}{\tablename~\oldnumberline}%
\listoftables%
}
\newpage
\phantomsection
\addcontentsline{toc}{section}{List of \figurename{s}}
{% See comment on list of tables
\let\oldnumberline\numberline%
\renewcommand{\numberline}{\figurename~\oldnumberline}%
\listoffigures%
}
\newpage

% make bookmarks for figure and tables sections
\renewcommand{\figuresection}{\pdfbookmarkraised{0}{\figurename{s}}{figure}\figurename{s}}
\renewcommand{\tablesection}{\pdfbookmarkraised{0}{\tablename{s}}{table}\tablename{s}}

\pagenumbering{arabic}
\section{Background} \label{sec:background}

@@

\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:workflow}: Batch adjustment justification and steps}{fig:workflow}
	\centering
	% TODO I changed the width so "Figures" would be on the same page as the 1st fig per Gentri's request. Change this back eventually
	\includegraphics[width=0.85\columnwidth]{figures/final/adjuster_workflow.pdf}
	\caption[Batch adjustment justification and steps]{\textbf{Batch adjustment justification and steps.}
	\begin{enumerate*}[(a)]
		\item When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
		\item If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
		\item In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
		\item Second, we remove the estimated effects in order to normalize the batches relative to one another.
	\end{enumerate*}
	}
	\label{fig:workflow}
\end{figure}

@@

All our code has been made publicly available at \nolinkurl{https://github.com/jdayton3/Confounded}, and all our data are available as described below.

@@

\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:network}: Network architecture of Confounded}{fig:network}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
	\caption[Network architecture of Confounded]{\textbf{Network architecture of Confounded.}
	Data with batch effects (represented by different colors) are input into an autoencoder.
	The output of the autoencoder is classified by a discriminator network based on batch.
	The autoencoder is then penalized based on the success of the discriminator.
	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
	\label{fig:network}
\end{figure}

@@

\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:training_loss}: Autoencoder and discriminator loss}{fig:training_loss}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/training_loss.pdf}
	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded on the MNIST dataset.
	Over the course of training, the autoencoder more faithfully replicates the input data.
	The autoencoder also seems to introduce noise (see the red dashed line around iteration 3100) in response to the discriminator's slight improvements.}
	\label{fig:training_loss}
\end{figure}

@@

\begin{table}
	\centering
	\input{figures/final/datasets.tex}
	\caption[Dataset information]{
		\pdfbookmarkraised{1}{Table \ref{tab:datasets}: Dataset information}{tab:datasets}
		\textbf{Dataset information} for each dataset used.
	}
	\label{tab:datasets}
\end{table}

@@

% \paragraph{Synthetic Batch Effects} % TODO: uncomment this after submitting thesis

@@

The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
We accessed a version of this dataset from \nolinkurl{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
We used this cancer status variable as the ``true class'' for classification.
It has two types of batch variables: ``plate'' and ``centre.''
We adjusted against the ``plate'' variable because it was more balanced than ``centre'' (with counts of $\{43, 50\}$ compared to $\{27, 66\}$).
The GSE37199 dataset represents a slightly larger dataset than bladderbatch, with only two batches that are closer to being balanced (with batch counts of $\{4, 5, 11, 18, 19\}$ and $\{43, 50\}$, respectively).

@@

The Cancer Genome Atlas (TCGA) Pan-Cancer project produced expression data for thousands of tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
In a previous study \cite{dayton_classifying_2017-1}, we classified this dataset based on the presence or absence of mutations in several known cancer genes.
We attempted to adjust for the confounding effect of cancer type using ComBat prior to classification.
However, we found that a strong nonlinear signal could still be identified by the Random Forests algorithm after adjustment.
Here, we used the same version of the dataset that we tidied in this previous study (available at \nolinkurl{https://osf.io/7xjdn/}).
This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
We used this dataset as a way to test whether Confounded works on RNA-Seq data and to test whether we could remove confounding effects that ComBat cannot remove.

@@

We compared our method to two other batch adjusters: a scale adjuster and ComBat \citep{johnson_adjusting_2007}.
We implemented the scale adjuster in the R programming language, version 3.6.0 \citep{r_core_team_r_2014} using RStudio version 1.2.1194 \citep{rstudio_team_rstudio_2018}.
It adjusts the data by linearly expanding or contracting each batch so all batches have the same range.
We used the ComBat implementation from the R sva package \citep{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
We initially intended to test against the SVA \citep{leek_capturing_2007} method but concluded that SVA is more suited for producing surrogate variables for further statistical research rather than removing those variables from the data.
There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \citep{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}). % TODO: add like 8 more methods
Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.
For these reasons, we compared Confounded only to the two methods listed above.
Future batch adjustment research may benefit from standardization of input formats, user interfaces, and validation datasets.

@@

\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:pca}: Principal components analysis (PCA)}{fig:pca}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/pca.pdf}
	\caption[Principal components analysis (PCA)]{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
	None of the datasets appear to be linearly separable.
	Confounded appears to maintain the same distribution of data overall as the unadjusted data while perhaps aligning the batches' distributions.}
	\label{fig:pca}
\end{figure}
\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:tsne}: T-distributed Stochastic Neighbor Embedding (t-SNE)}{fig:tsne}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/tsne.pdf}
	\caption[T-distributed Stochastic Neighbor Embedding (t-SNE)]{\textbf{T-distributed Stochastic Neighbor Embedding (t-SNE)} plot for the GSE37199 dataset before and after adjustment with several algorithms.
	The data seem to cluster less by batch for both Confounded and ComBat, indicating that both adjusters may be removing nonlinear effects in this dataset.}
	\label{fig:tsne}
\end{figure}

@@

\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:mse}: Mean squared error (MSE)}{fig:mse}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
	\caption[Mean squared error (MSE)]{
		\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
		Lower MSE represents that the adjuster has more faithfully reproduced the input data.
		MSE for unadjusted data will always be 0 because the input data is identical to the output data.
		Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
		See also Table \ref{tab:mse}.
	}
	\label{fig:mse}
\end{figure}
\begin{table}
	\input{figures/final/mse.tex}
	\caption[Mean squared error (MSE)]{
		\pdfbookmarkraised{1}{Table \ref{tab:mse}: Mean squared error (MSE)}{tab:mse}
		\textbf{Mean squared error (MSE)} of the unadjusted input data compared to the data output by the given adjusters.
		Lower MSE indicates that the output has changed less from the input.
		See also \figurename{} \ref{fig:mse}.
	}
	\label{tab:mse}
\end{table}
\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:mmd}: Maximum mean discrepancy (MMD)}{fig:mmd}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
	\caption[Maximum mean discrepancy (MMD)]{
		\textbf{Maximum mean discrepancy} (MMD) between different batches.
		Lower MMD indicates that the distributions of the different batches are more similar.
		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
		In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
		See also Table \ref{tab:mmd}.
	}
	\label{fig:mmd}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/mmd.tex}
	\caption[Maximum mean discrepancy (MMD)]{
		\pdfbookmarkraised{1}{Table \ref{tab:mmd}: Maximum mean discrepancy (MMD)}{tab:mmd}
		\textbf{Maximum mean discrepancy} (MMD) comparing the distributions of the batches to each other after a given adjustment.
		Lower MMD indicates that the distributions of the different batches are more similar.
		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
		See also \figurename{} \ref{fig:mmd}.
	}
	\label{tab:mmd}
\end{table}

@@

\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:batch}: Batch classification accuracy}{fig:batch}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/batch.tex}
	\caption[Batch classification accuracy]{
		\pdfbookmarkraised{1}{Table \ref{tab:batch}: Batch classification accuracy}{tab:batch}
		\textbf{Batch classification accuracy} for several datasets and adjusters.
		The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
		See also \figurename{} \ref{fig:batch}.
	}
	\label{tab:batch}
\end{table}

@@

\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:true_class}: True class classification accuracy}{fig:true_class}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times. A higher accuracy after adjustment is desired because it represents that the adjuster has not destroyed the true class signal.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption[True class classification accuracy]{
		\pdfbookmarkraised{1}{Table \ref{tab:true_class}: True class classification accuracy}{tab:true_class}
		\textbf{True class classification accuracy} for several datasets and adjusters.
		After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
		See also \figurename{} \ref{fig:true_class}.
	}
	\label{tab:true_class}
\end{table}

\begin{figure}
	\pdfbookmarkraised{1}{\figurename{} \ref{fig:mnist}: MNIST handwritten digits}{fig:mnist}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mnist.png}
	\caption[MNIST handwritten digits]{\textbf{MNIST handwritten digits}
	\begin{enumerate*}[(a)]
		\item before any adjustment,
		\item with artificial noise added,
		\item adjusted for noise by the scale adjuster,
		\item adjusted for noise by ComBat, and
		\item adjusted for noise by Confounded.
	\end{enumerate*}
	Although Confounded seems to remove more noise from the background, it struggles in some cases to accurately replicate the input data.}
	\label{fig:mnist}
\end{figure}

>>>>

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{authblk}
\usepackage{url}
\usepackage[inline,shortlabels]{enumitem}
\usepackage[square,numbers]{natbib}
\usepackage{floatrow}
\floatsetup[table]{capposition=top}
\usepackage[nomarkers,nolists]{endfloat}
\usepackage[margin=1in,letterpaper]{geometry}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage[hidelinks]{hyperref}
\usepackage{lineno}
\linenumbers

@@

\title{Adversarial Deep Neural Networks Effectively Remove Nonlinear Batch Effects from Gene-Expression Data}
\author{Jonathan Bryan Dayton}
\author[1]{Jonathan B. Dayton}
\author[1]{Stephen R. Piccolo}
\affil[1]{Department of Biology, Brigham Young University, Provo, UT 84602 USA}

\maketitle

\begin{abstract}
	Gene-expression profiling enables researchers to quantify transcription levels in cells, thus providing insight into functional mechanisms of diseases and other biological processes.
	However, because of the high dimensionality of these data and the sensitivity of measuring equipment, expression data often contains unwanted confounding effects that can skew analysis.
	For example, collecting data in multiple runs causes nontrivial differences in the data (known as batch effects),
	known covariates that are not of interest to the study may have strong effects,
	and there may be large systemic effects when integrating multiple expression datasets.
	Additionally, many of these confounding effects represent higher-order interactions that may not be removable using existing techniques that identify linear patterns.
	We created Confounded to remove these effects from expression data.
	Confounded is an adversarial variational autoencoder that removes confounding effects while minimizing the amount of change to the input data.
	We tested the model on artificially constructed data and commonly used gene expression datasets and compared against other common batch adjustment algorithms.
	We also applied the model to remove cancer-type-specific signal from a pan-cancer expression dataset.
	Our software is publicly available at \url{https://github.com/jdayton3/Confounded}.
\end{abstract}

\doublespacing
\section{Background} \label{sec:background}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
	\caption[Batch adjustment justification and steps]{\textbf{Batch adjustment justification and steps.}
	\begin{enumerate*}[(a)]
		\item When measurements are collected from a sample ($X$), systemic effects ($H$) also affect the measurements.
		\item If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
		\item In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
		\item Second, we remove the estimated effects in order to normalize the batches relative to one another.
	\end{enumerate*}
	}
	\label{fig:workflow}
\end{figure}

@@

All our code has been made publicly available at \url{https://github.com/jdayton3/Confounded}, and all our data are available as described below.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
	\caption[Network architecture of Confounded]{\textbf{Network architecture of Confounded.}
	Data with batch effects (represented by different colors) are input into an autoencoder.
	The output of the autoencoder is classified by a discriminator network based on batch.
	The autoencoder is then penalized based on the success of the discriminator.
	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
	\label{fig:network}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/training_loss.pdf}
	\caption[Autoencoder and discriminator loss]{\textbf{Autoencoder and discriminator loss} over time for one run of Confounded on the MNIST dataset.
	Over the course of training, the autoencoder more faithfully replicates the input data.
	The autoencoder also seems to introduce noise (see the red dashed line around iteration 3100) in response to the discriminator's slight improvements.}
	\label{fig:training_loss}
\end{figure}

@@

\begin{table}
	\centering
	\input{figures/final/datasets.tex}
	\caption[Dataset information]{
		\textbf{Dataset information} for each dataset used.
	}
	\label{tab:datasets}
\end{table}

@@

\paragraph{Synthetic Batch Effects}

@@

The GSE37199 dataset contains Affymetrix microarray gene-expression data from patients with advanced castration-resistant prostate cancer \cite{olmos_prognostic_2012}.
We accessed a version of this dataset from \url{http://doi.org/10.17605/OSF.IO/SSK3T} that was tidied as part of a curated compendium of human transcriptional biomarker data \cite{golightly_curated_2018}.
It contains expression values for 93 tissue samples categorized as either ``Advanced castration resistant'' or ``good prognosis.''
We used this cancer status variable as the ``true class'' for classification.
It has two types of batch variables: ``plate'' and ``centre.''
We adjusted against the ``plate'' variable because it was more balanced than ``centre'' (with counts of $\{43, 50\}$ compared to $\{27, 66\}$).
The GSE37199 dataset represents a slightly larger dataset than bladderbatch, with only two batches that are closer to being balanced (with batch counts of $\{4, 5, 11, 18, 19\}$ and $\{43, 50\}$, respectively).

@@

The Cancer Genome Atlas (TCGA) Pan-Cancer project produced expression data for thousands of tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
In a previous study \cite{dayton_classifying_2017-1}, we classified this dataset based on the presence or absence of mutations in several known cancer genes.
We attempted to adjust for the confounding effect of cancer type using ComBat prior to classification.
However, we found that a strong nonlinear signal could still be identified by the Random Forests algorithm after adjustment.
Here, we used the same version of the dataset that we tidied in this previous study (available at \url{https://osf.io/7xjdn/}).
This dataset has RNA-Seq expression values for 9,365 samples across 25 distinct cancer types.
We used this dataset as a way to test whether Confounded works on RNA-Seq data and to test whether we could remove confounding effects that ComBat cannot remove.

@@

We compared our method to two other batch adjusters: a scale adjuster and ComBat \citep{johnson_adjusting_2007}.
We implemented the scale adjuster in the R programming language, version 3.6.0 \citep{r_core_team_r_2014} using RStudio version 1.2.1194 \citep{rstudio_team_rstudio_2018}.
It adjusts the data by linearly expanding or contracting each batch so all batches have the same range.
We used the ComBat implementation from the R sva package \citep{leek_sva_2017} with some modifications to allow it to work on columns without variance in the MNIST dataset.
We initially intended to test against the SVA \citep{leek_capturing_2007} method but concluded that SVA is more suited for producing surrogate variables for further statistical research rather than removing those variables from the data.
There are a number of other methods for batch adjustment that do and do not use deep neural networks (for example, see \citep{leek_capturing_2007,espin-perez_comparison_2018,shaham_removal_2017,shaham_batch_2018}).
Unfortunately, most of these methods lack a common interface and common assumptions that input datasets must meet.
For these reasons, we compared Confounded only to the two methods listed above.
Future batch adjustment research may benefit from standardization of input formats, user interfaces, and validation datasets.

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/pca.pdf}
	\caption[Principal components analysis (PCA)]{\textbf{Principal components analysis} (PCA) of the GSE37199 dataset before and after batch adjustment with various adjusters.
	None of the datasets appear to be linearly separable.
	Confounded appears to maintain the same distribution of data overall as the unadjusted data while perhaps aligning the batches' distributions.}
	\label{fig:pca}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/tsne.pdf}
	\caption[T-distributed Stochastic Neighbor Embedding (t-SNE)]{\textbf{T-distributed Stochastic Neighbor Embedding (t-SNE)} plot for the GSE37199 dataset before and after adjustment with several algorithms.
	The data seem to cluster less by batch for both Confounded and ComBat, indicating that both adjusters may be removing nonlinear effects in this dataset.}
	\label{fig:tsne}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
	\caption[Mean squared error (MSE)]{
		\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
		Lower MSE represents that the adjuster has more faithfully reproduced the input data.
		MSE for unadjusted data will always be 0 because the input data is identical to the output data.
		Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
		See also Table \ref{tab:mse}.
	}
	\label{fig:mse}
\end{figure}
\begin{table}
	\input{figures/final/mse.tex}
	\caption[Mean squared error (MSE)]{
		\textbf{Mean squared error (MSE)} of the unadjusted input data compared to the data output by the given adjusters.
		Lower MSE indicates that the output has changed less from the input.
		See also \figurename{} \ref{fig:mse}.
	}
	\label{tab:mse}
\end{table}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
	\caption[Maximum mean discrepancy (MMD)]{
		\textbf{Maximum mean discrepancy} (MMD) between different batches.
		Lower MMD indicates that the distributions of the different batches are more similar.
		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
		In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
		See also Table \ref{tab:mmd}.
	}
	\label{fig:mmd}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/mmd.tex}
	\caption[Maximum mean discrepancy (MMD)]{
		\textbf{Maximum mean discrepancy} (MMD) comparing the distributions of the batches to each other after a given adjustment.
		Lower MMD indicates that the distributions of the different batches are more similar.
		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
		See also \figurename{} \ref{fig:mmd}.
	}
	\label{tab:mmd}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/batch.tex}
	\caption[Batch classification accuracy]{
		\textbf{Batch classification accuracy} for several datasets and adjusters.
		The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
		See also \figurename{} \ref{fig:batch}.
	}
	\label{tab:batch}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times. A higher accuracy after adjustment is desired because it represents that the adjuster has not destroyed the true class signal.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption[True class classification accuracy]{
		\textbf{True class classification accuracy} for several datasets and adjusters.
		After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
		See also \figurename{} \ref{fig:true_class}.
	}
	\label{tab:true_class}
\end{table}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mnist.png}
	\caption[MNIST handwritten digits]{\textbf{MNIST handwritten digits}
	\begin{enumerate*}[(a)]
		\item before any adjustment,
		\item with artificial noise added,
		\item adjusted for noise by the scale adjuster,
		\item adjusted for noise by ComBat, and
		\item adjusted for noise by Confounded.
	\end{enumerate*}
	Although Confounded seems to remove more noise from the background, it struggles in some cases to accurately replicate the input data.}
	\label{fig:mnist}
\end{figure}
